Parity-odd Neutrino Torque Detection

### 2018-10-20 neu_torque.tex
title and abstract

%Title of paper
\title{Neutrino Torque}

@@

\begin{abstract}
Neutrino mass is a long-standing physics problem. The oscillation experiments and cosmological observations provide respectively lower and upper bound on their sum of mass. Using the cosmic large scale structures to constrain the neutrino mass we encounter difficulties in disentangling other degeneracy parameters and cosmic variance. Here we show that the presence of neutrino mass provides a unique contribution to spin directions of galaxies. This neutrino torque effect is unaffected by galaxy bias and cosmic variance. Recent isobaric reconstruction techniques can reliably predict the tidal shear field of neutrinos and the resulting neutrino torque can be predicted. Upcoming galaxy surveys are promising to use the neutrino torque effect to independently constrain the neutrino mass.

@@

\tcb{\textit{Introduction.---}}
The flavour oscillation experiments \citep{2002PhRvL..89a1301A} discovered the mass splittings of neutrinos and placed a lower bound of the sum of their mass $M_\nu \equiv \sum_{i=1}^3 m_{\nu_i} \gtrsim$ 0.05 eV \citep{2014ChPhC..38i0001O}. 
The existence of neutrino mass has profound impacts on cosmic evolution. 
In the early Universe relativistic neutrinos modulate the matter-to-radiation ratio, and based on which the cosmic microwave background observations provide an upper bound of $M_\nu\lesssim 0.23$ eV \citep{2016A&A...594A..13P}. 
In the late Universe, neutrinos become non-relativistic, and contribute to the matter energy density $\Omega_m$. 
Unlike the majority of matter, the cold dark matter (CDM) and baryons, neutrinos maintain a high velocity dispersion, known as ``free streaming'', which reduces their gravitational collapse on small scales. A number of large scale structure (LSS) surveys 
(such as Euclid \cite{2011arXiv1110.3193L},
LSST \cite{2009arXiv0912.0201L}, eBOSS \cite{2016AJ....151...44D},
WFIRST\cite{2012SPIE.8442E..1UG}, and DESI \cite{2015AAS...22533605E}) 
are aimed to improve this upper bound using neutrino effects on LSS,
however we usually encounter difficulties in disentangling neutrino effects from other parameters, and in understanding halo bias and cosmic variance.
Recently, new nonlinear neutrinos effects on LSS are proposed to give independent measurements on neutrino mass \citep{2014PhRvL.113m1301Z,2016PhRvL.116n1301Z,2017NatAs...1E.143Y}.
Here we present a novel neutrino torque effect which modulates the angular momenta (spin) of dark matter halos, which can be measured to constrain the neutrino mass.

\tcb{\textit{Theory.---}}
In the picture of LSS formation, gravitational instability lets initial density fluctuations form dark matter halos, in which galaxies are embedded. While in these highly nonlinear structures, uncertainties of halo bias, halo merging history and baryonic mechanisms obstruct us from clearly understanding the statistics like number counts and morphologies, the spins of galaxies/halos represent a clean structure formation history in the linear Universe, which are poorly appreciated.

@@

\tcb{Measured from simulations, $(\alpha_1,\alpha_2,\alpha_3)=(-0.18,0.18,0.02)$ optimize the cross-correlation $\left\langle\bs{j}^I_{\nu 0}\cdot\bs{j}^{I_r}_\nu\right\rangle=0.19$. In this case we need $4\times 10^8$ halos to have a $1\sigma$ detection (see Appendix A)}.

\tcb{\textit{Discussion.---}} 
We notice that the coefficients are prominently higher than \citep{2000ApJ...532L...5L}, and we investigate numerical errors that may affect the results. P3M (particle-particle particle-mesh) algorithms result in higher $\mu_L$ and $\mu_I$ compared to PM (particle-mesh), where additional tangential forces in PM violate the angular momentum conservation. Higher mass halos in a given simulation generally have slightly higher $\mu_L$ and $\mu_I$, however the correlation is enhanced greatly as we use higher mass resolutions.

@@

@@

% Specify following sections are appendices. Use \appendix* if there
% only one appendix.
\appendix
\section{How many halos do we need in 3D}
There are $N$ halos with their unit spin vector randomly distributed on a 2D sphere, $|\bs{j}|=1$ and $\left\langle \bs{j} \right\rangle =\bs{0}$. Adding an additional vector $a\hat{\bs{x}}$ ($a\ll 1$) to $\bs{j}$ and normalize we have $\bs{j}'=(\bs{j}+a\hat{\bs{x}})/|\bs{j}+a\hat{\bs{x}}|$,then we project $\bs{j}'$ onto $\hat{\bs{x}}$ and we get the signal $b=\bs{j}'\cdot\hat{\bs{x}}$. The numerical statistics are as following: $\left\langle b \right\rangle=2a/3$ and $\sigma(b)=1/\sqrt{3N}$. So, to get a $n\sigma$ detection, $N=3n^2/4a^2$ halos will be needed.

>>>>

%Title of paper
\title{Parity-odd Neutrino Torque Detection}

@@

\begin{abstract}
Cosmological observations are promising ways for discovering the neutrino mass properties. The upper bound on their sum of mass is given by the cosmic microwave background and large scale structures. These measurements are all parity-even. 
Here we show that, the presence of neutrino mass provides a unique contribution to the directions of the angular momentum  of galaxies, which is the first parity-odd neutrino effect on large scale structures. This parity-odd observable is free of the contamination of the linear perturbation theory, and can be cleanly separated from other non-gravitational effects. A complete 21-cm neutral Hydrogen halo survey deep to redshift 1 can give a $5\sigma$ confidence level of detecting the neutrino torque effect if the sum of neutrino masses is 0.05 eV.

@@

\tcb{\textit{Introduction.---}}
Neutrino mass is a long-standing physics problem. The flavour oscillation experiments \citep{2002PhRvL..89a1301A} discovered the mass splittings of neutrinos and placed a lower bound of the sum of their mass $M_\nu \equiv \sum_{i=1}^3 m_{\nu_i} \gtrsim$ 0.05 eV \citep{2014ChPhC..38i0001O}. 
The existence of neutrino mass has profound impacts on cosmic evolution. 
In the early Universe relativistic neutrinos modulate the matter-to-radiation ratio, and based on which the cosmic microwave background observations provide an upper bound of $M_\nu\lesssim 0.23$ eV \citep{2016A&A...594A..13P}. 
In the late Universe, neutrinos become non-relativistic, and contribute to the matter energy density $\Omega_m$. 
Unlike the majority of matter, the cold dark matter (CDM) and baryons, neutrinos maintain a high velocity dispersion, known as ``free streaming'', which reduces their gravitational collapse on small scales. A number of large scale structure (LSS) surveys 
(such as Euclid \cite{2011arXiv1110.3193L},
LSST \cite{2009arXiv0912.0201L}, eBOSS \cite{2016AJ....151...44D},
WFIRST\cite{2012SPIE.8442E..1UG}, and DESI \cite{2015AAS...22533605E}) 
are aimed to improve this upper bound using neutrino effects on LSS,
however we usually encounter difficulties in disentangling neutrino effects from other parameters, and in understanding halo bias and cosmic variance.
Recently, new nonlinear neutrinos effects on LSS are proposed to give independent measurements on neutrino mass \citep{2014PhRvL.113m1301Z,2016PhRvL.116n1301Z,2017NatAs...1E.143Y}.
Here we present a novel neutrino torque effect which modulates the angular momenta (spin) of dark matter halos, which can be measured to constrain the neutrino mass.

\tcb{\textit{Theory.---}} 
In the picture of LSS formation, gravitational instability lets initial density fluctuations form dark matter halos, in which galaxies are embedded. While in these highly nonlinear structures, uncertainties of halo bias, halo merging history and baryonic mechanisms obstruct us from clearly understanding the statistics like number counts and morphologies \citep{2018MNRAS.tmp.2447C}, the spins of galaxies/halos represent a clean structure formation history in the linear Universe, which are poorly appreciated.

@@

Measured from simulations, \tcb{$(\alpha_1,\alpha_2,\alpha_3)=(-0.18,0.18,0.02)$} optimizes the cross-correlation \tcb{$\left\langle\bs{j}^I_{\nu 0}\cdot\bs{j}^{I_r}_\nu\right\rangle=0.19$}. In this case we need \tcb{$4\times 10^8$} halos to have a $1\sigma$ detection (see Appendix A). If we improve the reconstruction of $\bold{I}_R$, the lower limit of required halos is \tcb{$1.7\times 10^7$}.

\tcb{\textit{Discussion.---}} 
Accuracy and mass dependency: we notice that the cross-correlation coefficients between $\bs{j}_0$ and initial values $\bs{j}_L$ and $\bs{j}_T$ are prominently higher than what was measured in \citep{2000ApJ...532L...5L}. We investigate that numerical errors that may affect the results. P3M (particle-particle particle-mesh) algorithms result in higher cross-correlation between initial and final spins, compared to PM (particle-mesh), where additional tangential forces in PM violate the angular momentum conservation. Higher mass halos in a given simulation generally have slightly higher $\mu_L$ and $\mu_I$, however the correlation is enhanced greatly as we use higher mass resolutions. All other cross-correlation measurements have only weak dependencies on the halo mass, even in a fixed simulation. With different box sizes, mass resolutions, force resolutions, and find that the results are consistent across these simulations. Especially, in the estimation of number of halo spins needed to detect the neutrino torque, the deciding number is the cross-correlation $\left\langle\bs{j}^I_{\nu 0}\cdot\bs{j}^{I_r}_\nu\right\rangle=0.19$. This has a very weak dependence on halo mass and configuration of the simulation.

Reconstruction: to extract the neutrino torque from observable galaxy surveys, we need to reconstruct the initial conditions. There are many emerging successful reconstruction algorithms \citep{2017PhRvD..96l3502Z,2017ApJ...847..110Y,2014ApJ...794...94W}, mostly based on the $E$-mode clustering in Lagrangian space \citep{2017PhRvD..95d3501Y}. Least action algorithms.

Baryonic effects (+ etc.) in reconstruction and angular momentum conservation:

Observability: surveys like the Hubble Sphere Hydrogen Survey (HSHS) \citep{2006astro.ph..6104P} can cover 1 billion galaxies.

Beyond standard models: it was also suggested that neutrino mass is generated by a gravitational $\theta$-term condensation as the source of dark energy \citep{2016PhRvD..93k3002D}.

Cosmic variance:

Bias:

@@

\tcb{\textit{Acknowledgments.---}}
We acknowledge funding from NSERC.

@@

% Specify following sections are appendices. Use \appendix* if there
% only one appendix.
\appendix
\section{Errors in 3D}
There are $N$ halos with their unit spin vector randomly distributed on a 2D sphere, $|\bs{j}|=1$ and $\left\langle \bs{j} \right\rangle =\bs{0}$. Adding an additional vector $a\hat{\bs{x}}$ ($a\ll 1$) to $\bs{j}$ and normalize we have $\bs{j}'=(\bs{j}+a\hat{\bs{x}})/|\bs{j}+a\hat{\bs{x}}|$, then we project $\bs{j}'$ onto $\hat{\bs{x}}$ and we get the signal $b=\bs{j}'\cdot\hat{\bs{x}}$. The numerical statistics are as following: $\left\langle b \right\rangle=2a/3$ and $\sigma(b)=1/\sqrt{3N}$. So, to get a $n\sigma$ detection, 
\begin{equation}
	N=3n^2/4a^2 
\end{equation}
halos will be needed.
